\bf Fast and Efficient Black Box Optimization using the Parameter-less Population Pyramid

### 2015-03-03 Fast and Efficient Black Box Optimization using the Parameter-less Population Pyramid/main.tex
LTGA and hBOA with local search.

\begin{figure}
  \begin{algorithmic}[1]
  \Procedure{Iterate-P3}{}
    \State Create random solution
    \State Apply hill climber
    \If{solution $\notin hashset$}
      \State Add solution to $P_0$ and rebuild model
      \State Add solution to $hashset$
    \EndIf

    \ForAll{$P_i \in pyramid$}
      \State Mix solution with $P_i$
      \If{solution's fitness has improved}
        \If{solution $\notin hashset$}
          \State Add solution to $P_{i+1}$ and rebuild model
          \State Add solution to $hashset$
        \EndIf
      \EndIf
    \EndFor
  \EndProcedure
\end{algorithmic}
  \caption{One iteration of P3 optimization. $pyramid$ is an
           ordered set of populations and $hashset$ is a set
           of all solutions in $pyramid$.}
  \label{fig-p3}
\end{figure}

@@

Figure~\ref{fig-pop-sizes} shows the results from performing bisection. In general hBOA
required population sizes which were at least an order of magnitude larger than LTGA. Due to
runtime and memory overhead, finding the optimal value for hBOA was much less tractable than LTGA
for moderate to large problem sizes. LTGA's population size also grew significantly slower
than hBOA's as problem size increased, especially on the two Trap problems and Rastrigin.
Both algorithms were very ineffective on MAX-SAT, neither able to tune to problems sizes over 60 bits.
This is likely due to the fact that some randomly generated MAX-SAT landscapes are very flat
and highly deceptive~\citep{rana:1998:gamaxsat}.
\section{Finding the Global Optimum}
\label{sec-optimum}

>>>>

\begin{figure}
  \begin{algorithmic}[1]
  \Procedure{Iterate-P3}{}
    \State Create random solution
    \State Apply hill climber
    \If{solution $\notin hashset$}
      \State Add solution to $P_0$ and rebuild $P_0$'s model
      \State Add solution to $hashset$
    \EndIf

    \ForAll{$P_i \in pyramid$}
      \State Mix solution with $P_i$
      \If{solution's fitness has improved}
        \If{solution $\notin hashset$}
          \State Add solution to $P_{i+1}$ and rebuild $P_{i+1}$'s model
          \State Add solution to $hashset$
        \EndIf
      \EndIf
    \EndFor
  \EndProcedure
\end{algorithmic}
  \caption{One iteration of P3 optimization. $pyramid$ is an
           ordered set of populations and $hashset$ is a set
           of all solutions in $pyramid$.}
  \label{fig-p3}
\end{figure}

@@

Figure~\ref{fig-pop-sizes} shows the results from performing bisection. In general hBOA
required population sizes which were at least an order of magnitude larger than LTGA. Due to
runtime and memory overhead, finding the optimal value for hBOA was much less tractable than LTGA
for moderate to large problem sizes. LTGA's population size also grew significantly slower
than hBOA's as problem size increased, especially on the two Trap problems and Rastrigin.
Both algorithms were very ineffective on MAX-SAT, neither able to tune to problems sizes over 60 bits.
This is likely due to the fact that some randomly generated MAX-SAT landscapes are very flat
and highly deceptive~\citep{rana:1998:gamaxsat}.

While not currently treated as a parameter, we also performed preliminary tests of
integrating hill climbing into LTGA and hBOA. To match P3, we applied first improvement
hill climbing to each algorithm's initial population.
We then performed bisection on the modified algorithms for the largest problem sizes
where hBOA without hill climbing was effective. We found that in general both methods
performed worse when combined with hill climbing, in some cases up to an order of
magnitude worse. There were three exceptions: both improved on MAX-SAT and
hBOA improved on Rastrigin. In all cases the inclusion of hill climbing did
not result in either algorithm outperforming P3 in terms of evaluations
required to reach the global optimum. As such, all further experiments
use the unmodified published versions.

\section{Finding the Global Optimum}
\label{sec-optimum}
